---
title: "graduation.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "lumen"
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [graduation.csv](https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/graduation.csv) include student-level attributes for $n=2,344$ randomly sampled students who were first-year, full-time students from the 2002 cohort at a large, midwestern research university. Any students who transferred to another institution were removed from the data. The source of these data is @Jones-White:2014. The attributes, collected for these students are:

- `student`: Student ID number in the dataset
- `degree`: Did the student obtain a degree (i.e., graduate) from the institution? (No; Yes)
- `act`: Student's ACT score (If the student reported a SAT score, a concordance table was used to transform the score to a comparable ACT score.)
- `scholarship`: Amount of scholarship offered to student (in thousands of dollars)
- `ap_courses`: Number of Advanced Placement credits at time of enrollment
- `first_gen`: Is the student a first generation college student? (No; Yes)
- `non_traditional`: Is the student a non-traditional student (older than 19 years old at the time of enrollment)? (No; Yes)



### Preview

```{r}
# Import data
graduation = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/graduation.csv")

# View data
graduation
```


### References



