---
title: "mammals.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "lumen"
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The data in [mammals.csv](https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/mammals.csv) come from @Allison:1976 and contain data on 62 species of mammals. The attributes include:

- `species`: Common name of species
- `body_weight`: Body weight, in *kg*
- `brain_weight`: Brain weight, in *g*
- `slow_wave`: Average slow wave (non-dreaming) sleep per day, in hours
- `paradox`: Average paradoxical (dreaming) sleep per day, in hours
- `total_sleep`: Average sleep per day, in hours
- `lifespan`: Average lifespan, in years
- `gestation`: Average gestation period, in days
- `predation`: Rating of degree to which species are preyed upon, 1 (low)--5 (high)
- `exposure`: Rating of sleep exposure from 1 (low exposure; e.g., sleep in a burrow or den)--5 (maximally exposed sleep)
- `danger`: Rating of predatory danger based on predation and exposure scales, 1 (low danger)--5 (high degreee of danger)
					
									


### Preview

```{r}
# Import data
mammals = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/mammals.csv")

# View data
mammals
```


### References


