---
title: "minneapolis.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "lumen"
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [minneapolis.csv](https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/minneapolis.csv) were provided in @Long:2012. They constitute a sample of $n=22$ students taken from a much larger dataset collected by the Minneapolis Public School District. The data were collected to comply with the No Child Left Behind Act of 2001 and began during the 2004-05 school year. The variables included in the data are:

- `student_id`: De-identified student ID number
- `reading_score`: Reading achievement score$^\dagger$. 
- `grade`: Grade-level.
- `special_ed`: Is the student recieving special education services?
- `attendance`: Proportion of attendance based on the number of school days the student attended school during the four year of the study.


$^\dagger$The reading achievement scores are based on the reading section of th Northwest Achievement Levels Test (NALT), a multiple-choice, adaptive assessment of students' academic achievement. The NALT raw scores were converted to vertically equated scaled scores using an IRT model. Higher scale scores indicate more reading achievement.


### Preview

```{r}
# Import data
mpls = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/minneapolis.csv")

# View data
mpls
```


### References




