---
title: "movies.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    df_print: paged
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print = "75")

# knitr options
opts_chunk$set(
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
  )
```


The data in [movies.csv](https://github.com/zief0002/epsy-8252/raw/master/data/movies.csv) includes attributes for $n=292$ movies. These data are a subset of data from the `movies` data object included in the **ggplot2movies** package. The original data contains information on 24 variables collected from 28,819 movies. The attributes in this sample include::

- `title`: Movie's title
- `budget`: Movie's budget (in millions of U.S. dollars); adjusted to 2018 dollars
- `year`: Release date of the movie
- `age`: Age of the movie; Computed by subtracting the movie's release date from 2019
- `genre`: Movie genre (Action, Comedy, Drama, Other)
- `length`: Running time of the movie (in minutes)


### Preview

```{r}
# Import data
movies = readr::read_csv(file = "https://github.com/zief0002/epsy-8252/raw/master/data/movies.csv")

# View data
movies
```


### References

Wickham, H. (2015). *ggplot2movies: Movies data.* R package version 0.0.1. https://CRAN.R-project.org/package=ggplot2movies
