---
title: "nels.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    df_print: paged
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print = "75")

# knitr options
opts_chunk$set(
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
  )
```


This data in [nels.csv](https://github.com/zief0002/epsy-8252/raw/master/data/nels.csv) includes three attributes on $n = 744$ 10th-grade students. These data come from Keith (2015). The attributes are:

- `social10`: 10th-grade standardized achievement score on History/Civics/Georgraphy (Social Studies) test
- `self_esteem`: Standardized measure of self-esteem; administered in 10th-grade
- `locus`: Standardized measure of locus-of-control; more positive values indicate higher internal locus-of-control
- `ses`: Standardized measure of socio-economic status, including parent education level
- `gpa`: Grade-point average (GPA) from grades6&ndash;8 in all subjects (on a standard 4-point scale)
- `parent_ed`: Education-level (in years of schooling) for the parent with the highest level of education


### Preview

```{r}
# Import data
nels = readr::read_csv(file = "~/Documents/github/epsy-8252/data/nels.csv")

# View data
nels
```


### References

Keith, T. V. (2015). *Multiple regression and beyond: An introduction to multiple regression and structural equation modeling.* Routledge.

