---
title: "same-sex-marriage.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "lumen"
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```

The data in [same-sex-marriage.csv](https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/same-sex-marriage.csv) were collected from the 2008 American National Election Study, conducted jointly by the University of Michigan and Stanford University. These particular data consist of 1,746 American's responses. The attributes are:

- `support`: Dummy-coded variable indicating whether the respondent supports gay marriage? (1=Yes; 0=No)
- `attendance`: Frequency the respondent attends religious services (0=Never; 1=Few times a year; 2=Once or twice a month; 3=Almost every week; 4=Every week) 
- `denomination`: Respondent's religious denomination? (Catholic; Jewish; Protestant; Other)
- `friends`: Does the respondent have family or friends that are LGBT? (1=Yes; 0=No)
- `age`: Respondent's age, in years
- `female`: Dummy-coded variable indicating whether the respondent is female (1=Yes; 0=No)


### Preview

```{r}
# Import data
same_sex = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/same-sex-marriage.csv")

# View data
same_sex
```


<!-- ### References -->
