---
title: "usnews.csv"
output: 
  rmdformats::html_docco:
    css: 'codebooks.css'
    df_print: paged
---


```{r knitr_init, echo=FALSE, cache=FALSE}
library(knitr)
library(rmdformats)
library(fontawesome)

## Global options
options(max.print = "75")

# knitr options
opts_chunk$set(
  prompt = FALSE,
  tidy = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
  )
```


The data in [usnews.csv](https://raw.githubusercontent.com/zief0002/epsy-8252/master/data/usnews.csv) includes 13 attributes collected from the $n=129$ graduate schools of education ranked in the *U.S. News and World Report's 2018 Best Graduate Schools*. The attributes are:

- `rank`: Rank in USNWR
- `school`: Graduate program of Education
- `score`: Overall score given by USNWR
- `peer`: Peer assessment score (5.0 = highest)
- `expert_score`: Administrator/expert assessment score (5.0 = highest)
- `gre_verbal`: Mean GRE verbal score in 2016
- `gre_quant`: Mean GRE quantitative score in 2016
- `doc_accept`: Acceptance rate for doctoral students in 2016
- `student_faculty_ratio`: Ratio of doctoral students to faculty members in 2016
- `phd_granted_per_faculty`: Doctorates granted per faculty member in 2015–16
- `funded_research`: Funded research (in millions of dollars)
- `funded_research_per_faculty`: Funded research per faculty member (in thousands of dollars)
- `enroll`: Total graduate education enrollment in 2016


### Preview

```{r}
# Import data
usnews = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-8252/master/data/usnews.csv")

# View data
usnews
```


<!-- ### References -->


