---
title: "vocabulary.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "lumen"
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The data in [vocabulary.csv](https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/vocabulary.csv), adapted from data provided by @Bock:1975, come from the Laboratory School of the University of Chicago and include scaled test scores across four grades from the vocabulary section of the Cooperative Reading Test for $n=64$ students. The attributes in the dataset include:

- `id`: The student ID number
- `vocab_08`: The scaled vocabulary test score in 8th grade
- `vocab_09`: The scaled vocabulary test score in 9th grade
- `vocab_10`: The scaled vocabulary test score in 10th grade
- `vocab_11`: The scaled vocabulary test score in 11th grade
- `female`: Dummy coded sex variable (0 = Not female, 1 = Female)


### Preview

```{r}
# Import data
vocab = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/vocabulary.csv")

# View data
vocab
```


### References


