---
title: "wine.csv"
output: 
  html:
    css: '../assets/codebooks.css'
    df_print: paged
    code_folding: show
theme: "lumen"
highlight-style: "atom-one"
bibliography: '../assets/epsy8252.bib'
csl: '../assets/apa-single-spaced.csl'
---


```{r}
#| include: false
source("../assets/codebook-setup.R")
```


The data in [wine.csv](https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/wine.csv) includes data on 200 different wines. These data are a subset of a larger database ($n = 6,613$) from [wine.com](https://www.wine.com/), one of the biggest e-commerce wine retailers in the U.S. It allows customers to buy wine according to any price range, grape variety, country of origin, etc. The data were made available at [http://insightmine.com/](http://insightmine.com/bring-your-own-data-analyzing-wine-market/). The attributes include:

- `wine`: Wine name
- `vintage`: Year the wine was produced (centered so that 0 = 2008, 1 = 2009, etc.) 
- `region`: Region of the world where the wine was produced. These data include seven regions (Australia, California, France, Italy, New Zealand, South Africa, South America).
- `varietal`: Grape varietal These data include nine varietals (Cabernet Sauvignon, Chardonnay, Merlot, Pinot Noir, Sauvignon Blanc, Syrah/Shiraz, Zinfandel, Other Red, Other Whites).
- `rating`: Wine rating on a 100-pt. scale (these are from sources such as *Wine Spectator*, the *Wine Advocate*, and the *Wine Enthusiast*)
- `price`: Price in U.S. dollars


### Preview

```{r}
# Import data
wine = readr::read_csv(file = "https://raw.githubusercontent.com/zief0002/bespectacled-antelope/main/data/wine.csv")

# View data
wine
```


<!-- ### References -->
