---
title: "#protip"
subtitle: "Formatting p-Values that are Below .001 in Tables"
author: "Andrew Zieffler"
date: "2/11/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import Packages and Data



```{r import_data, message=FALSE, warning=FALSE}
# Load libraries
library(knitr)
library(kableExtra)
library(lmtest)
library(scales)
library(tidyverse)

# Read in data
mn = read_csv(file = "https://raw.githubusercontent.com/zief0002/epsy-8252/master/data/mn-schools.csv")
```


<br /><br />

## Likelihood Ratio Test

Here we will fit three models and perform the likelihood ratio test.

```{r}
# Fit models
lm.0 = lm(grad ~ 1, data = mn)
lm.1 = lm(grad ~ 1 + sat, data = mn)
lm.2 = lm(grad ~ 1 + sat + public, data = mn)

# LRT
lrtest(lm.0, lm.1, lm.2)
```

<br /><br />

We can use the `class()` function to understand how the output of the `lrtest()` is formatted.

```{r}
# Assign output into lrt
lrt = lrtest(lm.0, lm.1, lm.2)

# Check output class
class(lrt)
```

<br /><br />

The `lrtest()` function creates an `anova` object and, more importantly, a `data.frame` object. Since one of the outputs is a data frame, this means we can immediately pipe the `lrtest()` output into a `kable()` function in our RMD document.

```{r}
# Pipe into kable
lrt %>%
  kable()
```

<br /><br />

# Formatting the Table

The first thing I will do is rename the columns from the `lrtest()` output (before I pipe the results to `kable()`).

```{r}
# Get column names
names(lrt)

# Rename columns
names(lrt) = c("Parameters", "ln(Lik)", "df", "$\\chi^2$", "p")

lrt %>%
  kable()
```

<br /><br />

Now that the columns have more reasonable names, we can use the `pvalue()` function from the `{scales}` package to format the `p` column. This function includes a default argument `accuracy = 0.001` that will re-label any *p*-value that is less than .001 to `<.001`. To use it, we recreate the `p` column using `mutate()` to call the `pvalue()` function on the original `p` column.

```{r}
# Format p-values
lrt %>%
  mutate(p = pvalue(p)) %>%
  kable()
```

<br /><br />

Now you can use the `kable()` arguments and functions from `{kableExtra}` to spruce up the table.

```{r final_table}
# Set option so that any cell with 'NA' is left blank in the table
options(knitr.kable.NA = '')

# Create table
lrt %>%
  mutate(p = pvalue(p)) %>%
  add_column(Model = c("Model 1", "Model 2", "Model 3"), .before = 1) %>% #Include a model column before column 1
  kable(
    caption = "Results from a likelihood ratio test comparing three candidate models.",
    format = "html",
    escape = FALSE,
    align = "c",
    digits = c(NA, 0, 2, 0, 2, NA), #Note that all character columns can't be rounded, thus NA
    col.names = c("Model", "Parameters", "ln(Lik)", "<i>df</i>", "$\\chi^2$", "<i>p</i>"), #Italicize statistics in columns
    table.attr = "style='width:60%;'"
  )  %>%
  kable_classic() 
```


<br /><br />